Staged Abstract Interpreters

### 2019-08-15 paper/10_intro.tex
update intro

Statically analyzing semantic properties of a program is a
widely-known undecidable problem. Abstract interpretation as a
lattice-based approach to sound static analyses was proposed by
\citet{DBLP:conf/popl/CousotC77}. Equipped with Galois connections,
the analyzer can obtain the program runtime behaviors approximately by
computing the fixed points on the abstract domain. Despite the
tremendous theoretical development of abstract interpretation over the
years, constructing artifacts and analyzers that perform sound
abstract interpretation for modern and expressive languages was
considered abstruse and complicated for a long time.

In recent years, we observe rich progress on the methodologies for
constructing abstract interpreters from systematic principles rather than
ad-hoc engineering. Of particular note is the Abstracting Abstract
Machine (AAM) methodology \cite{DBLP:journals/jfp/HornM12,
  DBLP:conf/icfp/HornM10}. This methodology uncovers an approach to derive sound
abstract interpreters from their concrete counterparts, for example,
abstract machines, where the soundness can be easily established by
examining the transformation of semantic artifacts. For example, the
CEK machine \cite{DBLP:conf/popl/FelleisenF87} for concrete execution
can be readily refactored to an effective $0$-CFA control-flow
analysis \cite{Shivers:1988:CFA:53990.54007,
  Midtgaard:2012:CAF:2187671.2187672}: first, tweak the environment
dereference as a nondeterministic choice, such that the environment
may contain multiple possible values for a variable, then allocate
continuations in the environment to exploit return-flow information,
and also constrain the addresses space to be finite.  This systematic
abstraction approach can be tailored to different language features
(such as states, first-class controls, exceptions, and concurrency) and
sensitivity analyses \cite{DBLP:conf/icfp/Gilray0M16,
  DBLP:conf/popl/GilrayL0MH16, Darais:2015:GTM:2814270.2814308}. It
has also been applied to various small-step abstract machines
\cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10,
  Sergey:2013:MAI:2491956.2491979} and big-step definitional
interpreters \cite{Wei:2018:RAA:3243631.3236800,
  DBLP:journals/pacmpl/DaraisLNH17, Keidel:2018:CSP:3243631.3236767}.

Based on the idea of abstracting abstract machines, more
pragmatically, several implementation strategies utilizing purely
functional programming to build abstract interpreters have
emerged. Such techniques include monads, monad transformers
\cite{DBLP:journals/pacmpl/DaraisLNH17, Sergey:2013:MAI:2491956.2491979},
arrows \cite{Keidel:2018:CSP:3243631.3236767}, extensible effects
\cite{Githubsemantic}, etc.  The pure approaches provide certain
benefits, including that the abstract interpretation artifacts can be built
compositionally and modularly, e.g., by using monad
transformers. Therefore, the soundness of analysis can be proven with
less effort, either by mechanized \cite{Darais:2016:CGC:2951913.2951934}
or paper-based proofs \cite{Keidel:2018:CSP:3243631.3236767}. Also,
referential transparent and equational reasoning allow programmers to
reason the correctness of their implementations more confidently.

However, besides the intrinsic complexity of static analysis, there
are additional abstraction penalties with these high-level
implementation approaches.  First, similar to concrete interpreters,
the abstract interpreter analyzes the program by traversing the
abstract syntax tree (AST), which poses an interpretive overhead, such as
the pattern matching on the ASTs and recursive calls on the sub-expressions.
Such overhead can be negligible if the abstract
interpreter only runs on the program a single time, but can also
accumulate significantly if run repeatedly (for example, on
libraries).  Second, the abstract interpreter written in pure languages
usually extensively uses effect systems to implement the semantics of
abstract interpretation.  For example, the abstract interpreter that
returns all possible runtime values is a form of nondeterminism, where
nondeterminism monads can help.  Although such pure approaches have
their own merits and elegance, compared with imperative stateful
implementations, they are significantly slower.

In this paper, we propose an abstraction-without-regret approach to
eliminating performance penalties for abstract interpreters,
while still keeping the benefits inherent in purely functional
programming.  In short, our approach applies ideas and techniques from
program specialization and embedded DSLs to abstract interpreters. 1)
Using multi-stage programming , we can specialize the abstract
interpreter with respect to an input program and then generate
efficient low-level code that does the actual analysis.  The result of
specialization is reusable, and the effect layers have been eliminated
in the generated code. 2) Inspired by the tagless-final interpreters,
we define a generic interpreter that abstracts over binding-time and
different semantics, which allows user to implement different
semantics modularly, including the staged abstract interpretation
semantics.  Together with the multi-stage programming and type-level
annotations of binding-times, we can derive the staged abstract
interpreters without intrusive changes to the corresponding unstaged
one, thus the soundness is untouched.  In this sense, our approach is
no regret of both performance and engineering effort. We elaborate
these two main ideas in detail as follow.

@@

The idea of specializing interpreters can be traced to Futamura
projections \cite{Futamura1999, futamura1971partial}.  The first
Futamura projection specifically shows that specializing an
interpreter w.r.t. the input program yields an equivalent
executable. For instance, @eval@ is an interpreter for some language:
\begin{lstlisting}
  def eval(e: Expr)(arg: Input): Value
\end{lstlisting}
Given a program $e_0$ of type @Expr@, by applying the specialization,
we can obtain a specialized interpreter $\texttt{eval}_{\texttt{e0}}$
: @Input@ $\to$ @Value@, which is the so-called \textit{equivalent
  executable}.  By definition of the interpreter, they produce the
same result when applied with the argument
$\texttt{eval}_{\texttt{e0}}(arg) = [\![ e_0 ]\!] arg $.

Partial evaluation \cite{DBLP:books/daglib/0072559} was the first
proposed approach to realize Futamura projections: it first discovers
the binding-times (\textit{static} or \textit{dynamic}) of variables,
then evaluates the static part of the program, and generates a
residualized program that solely relies on the dynamic part. However,
it is hard to precisely analyze binding-times given an arbitrary
program.  As an alternative approach to specialization, multi-stage
programming (MSP) \cite{taha1999multi, DBLP:conf/pepm/TahaS97} lets
the programmer to explicitly control and annotate the binding-times in
the program, then the MSP system will check whether these annotations
are consistent and specialize the program using that information.  The
staging annotations can be either syntactic (e.g., quote and
quasi-quote in MetaML/MetaOCaml) or type-based (e.g., the Lightweight
Modular Staging framework \cite{DBLP:conf/gpce/RompfO10} in Scala).

Our proposed framework adopts the type-based multi-stage programming
from the Lightweight Modular Staging framework and implements the
Futamura projection of a big-step abstract interpreter for a small
stateful, higher-order language.  We use
\citet{DBLP:journals/pacmpl/DaraisLNH17}'s abstract definitional
interpreter that uses monad transformers as the unstaged baseline. By
deriving staged monads that can be used to do code generation, we may
obtain a staged abstract interpreter that fulfills the specialization.
After staging, the generated code is specialized to the input program,
and all monadic operations are all inlined and compiled down to
low-level Scala code.

@@

Although the four artifacts may look like very differently at first
glance, but in fact are all firmly rooted in the concrete semantics of
the language.  This observation provides a way to abstract over the
interpreter and achieve the flexibility of reinterpreting the shared
interpreter. We adapt the monadic interpreter schema, which already
have been applied to abstract interpreters, and introduce the
binding-time abstraction into the generic interpreter.  The generic
interpreter returns a value of monadic type, which can be varied by
different semantics. The domain of the interpreter and the effects
such as read, state and nondeterminism can be both wrapped into this
monadic type.  The binding-time abstraction is represented by a
higher-kinded type, and controls whether the interpreter produces
values directly or generates code.  It is worth to mention that the
binding-time type is also instrumented into the monadic type, so that
we will distinguish normal monads and staged monads.

>>>>

Abstract interpretation as a lattice-based approach to sound static
analyses was proposed by \citet{DBLP:conf/popl/CousotC77}. Equipped
with Galois connections \cite{CousotCousot79-1}, the analyzer can
soundly approximate the program runtime behaviors by computing fixed
points on the abstract domain. Despite the tremendous theoretical
development of abstract interpretation over the years, constructing
artifacts and analyzers that perform sound abstract interpretation for
modern and expressive languages was considered complicated for a long
time.

Recent progress on methodologies such as the Abstracting Abstract
Machine (AAM) \cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10}
uncovers an approach to construct abstract interpreters from
systematic principles rather than ad-hoc engineering.  By deriving
semantics artifacts for abstract interpretation from theire concrete
counterparts, for example, abstract machines, the soundness can be
easily established by examining the transformation of semantic
artifacts.
This systematic abstraction approach can be tailored to different
language features (such as state, first-class control, exception, and
concurrency) and sensitivity analyses \cite{DBLP:conf/icfp/Gilray0M16,
  DBLP:conf/popl/GilrayL0MH16, Darais:2015:GTM:2814270.2814308}. It
has also been applied to various small-step abstract machines
\cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10,
  Sergey:2013:MAI:2491956.2491979} and big-step definitional
interpreters \cite{Wei:2018:RAA:3243631.3236800,
  DBLP:journals/pacmpl/DaraisLNH17, Keidel:2018:CSP:3243631.3236767}.

Based on the idea of abstracting abstract machines, more
pragmatically, several implementation strategies utilizing purely
functional programming to build abstract interpreters have
emerged. Such techniques include monads, monad transformers
\cite{DBLP:journals/pacmpl/DaraisLNH17,
  Sergey:2013:MAI:2491956.2491979}, arrows
\cite{Keidel:2018:CSP:3243631.3236767}, extensible effects
\cite{Githubsemantic}, etc. The pure approaches provide certain
benefits, including that the abstract interpretation artifacts can be
built compositionally and modularly, e.g., by using monad
transformers. Therefore, the soundness of analysis can be proven with
less effort, by either mechanized proof
\cite{Darais:2016:CGC:2951913.2951934} or paper-based proofs
\cite{Keidel:2018:CSP:3243631.3236767}.  Also, referential
transparency and equational reasoning allow programmers to reason the
correctness of their implementations more confidently.

However, besides the intrinsic complexity of static analysis, there
are additional abstraction penalties with these high-level
implementation approaches. First, similar to concrete interpreters,
the abstract interpreter analyzes the program by traversing the
abstract syntax tree (AST), which poses an interpretive overhead, such
as the pattern matching on the ASTs and recursive calls on the
sub-expressions.
Such overhead can be negligible if the abstract interpreter only runs
on the program a single time, but can also accumulate significantly if
run repeatedly (for example, on libraries).  Second, it is common that
abstract interpreters written in pure languages extensively use effect
systems to implement the semantics of abstract interpretation. For
example, the abstract interpreter that returns all possible runtime
values is a form of nondeterminism, where nondeterminism monads can
help.  Although such pure approaches have their own merits, compared
with imperative stateful implementations, they are significantly
slower.

Roughly at the same when abstract interpretation was proposed in
1970s, \citet{futamura1971partial, Futamura1999} revealed a colse
connection between interpreters and compilers through a hierarchy of
specializations, namely, Futamura Projections. The first Futamura
projection shows that specializing an interpreter to the input program
removes interpretation overhead and yields an equivalent executable
that runs faster.  The challenge here is can we apply the same idea to
general abstract interpreters to make it run faster?
To achieve this, we need to handle a non-standard semantics that
extensively uses nondeterminism, operates on abstract domains, and
computes fixed-point iteratively with a cache.

In this paper, we show how to naturally extend the first Futamura
projection to abstract interpreters. We present an
abstraction-without-regret framework to eliminating performance
penalties for abstract interpreters, while keeping the benefits
inherent in purely functional programming. In short, our approach
borrows ideas and techniques from multi-stage programming and embedded
DSLs to abstract interpreters.  1) To remove the overhead from
interpretation and effect layers, we specialize the abstract
interpreter with respect to an input program and then generate
efficient low-level code that does the actual analysis.  2) Inspired
by the tagless-final interpreters
\cite{DBLP:journals/jfp/CaretteKS09}, we construct a generic
interpreter that abstracts over binding-time and different semantics,
which allows user to implement different semantics modularly,
including the staged abstract interpretation semantics. As the result,
we can derive the staged abstract interpreters without intrusive
changes to its corresponding unstaged one and preserve the soundness
of analysis. In this sense, our approach is no regret of both
performance and engineering effort. We elaborate these two main ideas
in detail as follows.

%The result of specialization is reusable, and the effect layers have
%been eliminated in the generated code.

@@

The first Futamura projection specifically shows that specializing an
interpreter w.r.t. the input program yields an equivalent
executable. For instance,
\lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{eval}
is an interpreter for some language:
\begin{lstlisting}
  def eval(e: Expr)(arg: Input): Value
\end{lstlisting}
Given a program $e_0$ of type
\lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{Expr},
by applying the specialization, we can obtain a specialized
interpreter $\texttt{eval}_{\texttt{e0}}$ :
\lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{Input}
$\to$
\lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{Value},
which is the so-called \textit{equivalent executable}.

% By definition of the interpreter, they produce the same result when applied with the argument $\texttt{eval}_{\texttt{e0}}(arg) = [\![ e_0 ]\!] arg $.

Partial evaluation \cite{DBLP:books/daglib/0072559} was the first
proposed approach to realize Futamura projections: it first discovers
the binding-times (\textit{static} or \textit{dynamic}) of variables,
then evaluates the static part of the program, and generates a
residualized program that solely relies on the dynamic part. However,
it is hard to precisely analyze binding-times given an arbitrary
program. As an alternative approach, multi-stage programming (MSP)
\cite{taha1999multi, DBLP:conf/pepm/TahaS97} lets users explicitly
programming over a two-level semantics \cite{NIELSON198859},
by manually annotating the binding-times of variables in the program.
Then the MSP system will check whether these annotations
are consistent and specialize the program using that information. The
staging annotations can be either syntactic (e.g., quote and
quasi-quote in MetaML/MetaOCaml) or type-based (e.g., the Lightweight
Modular Staging framework \cite{DBLP:conf/gpce/RompfO10} in Scala).

Similar to the two-level semantics \cite{NIELSON1989117}, we found
monads provide a good abstraction to hide the detail of abstract
interpretation semantics, and allows us to fit code generation into
the semantics. We use \citet{DBLP:journals/pacmpl/DaraisLNH17}'s
abstract definitional interpreter that uses monad transformers as the
unstaged baseline.  Our proposed framework adopts the type-based
multi-stage programming from the Lightweight Modular Staging framework
and implements the Futamura projection of a big-step abstract
interpreter for a small stateful, higher-order language. By deriving
staged monads that can be used to do code generation, we obtain a
staged abstract interpreter that fulfills the specialization. After
staging, the generated code is specialized to the input program, and
all monadic operations are all inlined and compiled down to low-level
Scala code.

@@

Although the four artifacts may look like very differently at first
glance, but in fact are all firmly rooted in the concrete semantics of
the language.  This observation provides a way to abstract over the
interpreter and achieve the flexibility of reinterpreting the shared
interpreter. Generic interpreters using monads have already been
applied to abstract interpreters, we further introduce the
binding-time abstraction into the generic interpreter.  The generic
interpreter returns a value of monadic type, which can be varied by
different semantics. The domain of the interpreter and the effects
such as read, state and nondeterminism can be both wrapped into this
monadic type.  The binding-time abstraction is represented by a
higher-kinded type, and controls whether the interpreter produces
values directly or generates code. It is worth to mention that the
binding-time type is also injected into the monadic type, so that we
will distinguish normal monads and staged monads.

### 2019-08-15 paper/40_absinterp.tex
update intro

@@

In out monad stack, the \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{in} and \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{out} cache modeled by a reader effect and
state effect, respectively. We first defined several methods to manipulate
the two caches through the monad stack; the implementations are elided.
\begin{lstlisting}
  def ask_in_cache: AnsM[Cache]; def get_out_cache: AnsM[Cache]
  def put_out_cache(out: R[Cache]): AnsM[Unit]
  def set_out_cache(cfg: R[Config], vs: R[(Value, Store)]): AnsM[Unit]
\end{lstlisting}

>>>>

\iffalse
For example, the CEK machine \cite{DBLP:conf/popl/FelleisenF87} for
concrete execution can be readily refactored to an effective $0$-CFA
control-flow analysis \cite{Shivers:1988:CFA:53990.54007,
  Midtgaard:2012:CAF:2187671.2187672}: first, tweak the environment
dereference as a nondeterministic choice, such that the environment
may contain multiple possible values for a variable, then allocate
continuations in the environment to exploit return-flow information,
and also constrain the address space to be finite.
\fi

@@

In our monad stack, the \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{in} and \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{out} cache modeled by a reader effect and
state effect, respectively. We first defined several methods to manipulate
the two caches through the monad stack; the implementations are elided.
\begin{lstlisting}
  def ask_in_cache: AnsM[Cache]; def get_out_cache: AnsM[Cache]
  def put_out_cache(out: R[Cache]): AnsM[Unit]
  def set_out_cache(cfg: R[Config], vs: R[(Value, Store)]): AnsM[Unit]
\end{lstlisting}
